---
title: "GSVA for mutil Group"
author: "Ximing Ran"
date: "2025-03-24"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: false
  html_document:
    # code_folding: hide
    toc: true
    toc_float: true
hitheme: tomorrow
highlighter: highlight.js

---

```{r setup, include=FALSE}
# load libraries
library(tidyverse)
library(knitr)
set.seed(2024)

knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.path = "./results/02-GSVA-4-group/Analysis_figure/"
)

theme_set(theme_bw() + theme(legend.position = "bottom"))
knitr::kable(head(mtcars[, 1:4]), "simple")
```

```{r}
library(tibble)
library(tidyr)
library(dplyr)
library(rtracklayer)
```



```{r local_function_load}
# load function from local files
source(here::here("source", "DEG_functions.R"))
```

\newpage

# 1. Read the count data
In this section, we will read the clean count data from the synaptosomes_bulkRNA folder.  We will read the data and merge them into a single table. 

```{r load_bulkRNA_data}
input_count <- read.csv(here::here("data", "bulkRNA",
                                      "bulkRNA_counts_cleaned.csv"))
counts <-  as.data.frame(input_count) %>% 
  column_to_rownames(var = "gene")
colnames(counts) <- gsub("_", "-", colnames(counts))


# raw sample list
sample_list_raw <- read.csv(here::here("data", "bulkRNA",
                                      "sample_info_AD.csv")) %>%
                    mutate(condition = paste0(Diagosis, "_", Treatment),
                           sample = gsub("_", "-", Sample.name))


# Ensure the column names of counts exist in Sample.name
new_colnames <- sample_list_raw$Label[match(colnames(counts), sample_list_raw$sample )]

# Assign new column names
colnames(counts) <- new_colnames


# sort the columns by the colname
condition_list <- data.frame(
  group =sample_list_raw$condition
)

row.names(condition_list) <- sample_list_raw$Label

counts<- counts[, rownames(condition_list)]

gene_name_mapping<- readRDS(here::here("data","ref" ,"gene_name_mapping.rds"))




```



```{r DESeq2_analysis}
# Init the result folder structure for the result
result_folder_all = './results'
result_folder = result_folder_all

```


\newpage

# 2. Visualization for reuslt

## (1) Sample information - PCA plot

```{r Sample_PCA, fig.width=8, fig.height=6}
figure_folder = result_folder
# do PCA for counts data
dds_obj <- DESeqDataSetFromMatrix(countData = counts,
                                  colData = condition_list,
                                  design = ~ group)
vsd.obj <- varianceStabilizingTransformation(dds_obj, blind = TRUE)
pcaData <- plotPCA(vsd.obj,  intgroup = c("group"), returnData = T)
percentVar <- round(100 * attr(pcaData, "percentVar"))


p <-ggplot(pcaData, aes(PC1, PC2, color=group)) +
  geom_point(size=3) +
  labs(x = paste0("PC1: ",percentVar[1],"% variance"),
       y = paste0("PC2: ",percentVar[2],"% variance"),
  ) +
  stat_ellipse(level = 0.95)+
  theme_bw() +
  # theme_classic()+
  theme(text = element_text(family = "Arial", colour = "black")) +
  # scale_color_manual(values = assigned_colors) +
  ggrepel::geom_text_repel(aes(label = name), color = "black")

print(p)
# ggsave("./results/01-Sample_info/01_sample_PCA_plot.pdf", p,width = 8, height = 6, units = "in", dpi = 300)
# ggsave("./results/01-Sample_info/01_sample_PCA_plot.png", p,width = 8, height = 6, units = "in", dpi = 300)
#   
```

## (2) Sample information - Distance heatmap

```{r Sample_dis_Heatmap, fig.width=8, fig.height=6}
 # Now apply variance stabilizing transformation
 sampleDists <- dist(t(assay(vsd.obj)))
 sampleDistMatrix <- as.matrix( sampleDists )
 rownames(sampleDistMatrix) <- paste( vsd.obj$group )
 colors <- colorRampPalette( rev(RColorBrewer::brewer.pal(9, "Blues")) )(255)
 p <- pheatmap::pheatmap(sampleDistMatrix,
                         clustering_distance_rows = sampleDists,
                         clustering_distance_cols = sampleDists,
                         col = colors) 
print(p)

#  ggsave("./results/01-Sample_info/02_sample_distance_heatmap.pdf", p,width = 8, height = 6, units = "in", dpi = 300)
# ggsave("./results/01-Sample_info/02_sample_distance_heatmap.png",
#        p, width = 8, height = 6, units = "in", dpi = 300)

```

\newpage



# 3. GSVA analysis

```{r GSVA_analysis, fig.width=12, fig.height=6}
# # The following code is used to generate the GSVA matrix , only need to run once
# gmxFile <- here::here("data", "ref", "c5.go.v2023.1.Hs.symbols.gmt")
# go_list <- getGmt(gmxFile)
# 
# geneset <- go_list
# dat <- as.matrix(counts)
# 
# gsvapar <- gsvaParam(dat, geneset, maxDiff=TRUE)
# gsva_es <- gsva(gsvapar)
# gsva_matrix <- as.data.frame(gsva_es)
# 
# # save the result
# write.csv(gsva_matrix, "./results/02-GSVA/01_GSVA_matrix.csv")
# 
# 



gsva_matrix_6 <- read.csv("./results/02-GSVA/01_GSVA_matrix.csv", row.names = 1)

condition_list_select <-condition_list %>% filter(group %in% c("CTRL_Veh_0", "AD_Veh_0",
                                                             "CTRL_Veh_10",   "AD_Veh_10"))

gsva_matrix_4 <- gsva_matrix_6[,rownames(condition_list_select)]

write.csv(gsva_matrix_4, "./results/02-GSVA/02_GSVA_matrix_4.csv")

# Get column names (sample IDs) for each group
ref_samples <- rownames(condition_list_select %>% filter(group == "CTRL_Veh_0"))
comp_samples <- rownames(condition_list_select %>% filter(group == "AD_Veh_0"))

# Perform row-wise t-test
gsva_matrix_4$CTRL_Veh_0_vs_AD_Veh_0_pvalue <- apply(gsva_matrix_4, 1, function(row) {
  t.test(as.numeric(row[ref_samples]),
         as.numeric(row[comp_samples]),
         paired = FALSE)$p.value
})


# Get column names (sample IDs) for each group
ref_samples <- rownames(condition_list_select %>% filter(group == "CTRL_Veh_10"))
comp_samples <- rownames(condition_list_select %>% filter(group == "AD_Veh_10"))

# Perform row-wise t-test
gsva_matrix_4$CTRL_Veh_10_vs_AD_Veh_10_pvalue <- apply(gsva_matrix_4, 1, function(row) {
  t.test(as.numeric(row[ref_samples]),
         as.numeric(row[comp_samples]),
         paired = FALSE)$p.value
})


# Get column names (sample IDs) for each group
ref_samples <- rownames(condition_list_select %>% filter(group == "CTRL_Veh_0"))
comp_samples <- rownames(condition_list_select %>% filter(group == "CTRL_Veh_10"))

# Perform row-wise t-test
gsva_matrix_4$CTRL_Veh_0_vs_CTRL_Veh_10_pvalue <- apply(gsva_matrix_4, 1, function(row) {
  t.test(as.numeric(row[ref_samples]),
         as.numeric(row[comp_samples]),
         paired = FALSE)$p.value
})


# Get column names (sample IDs) for each group
ref_samples <- rownames(condition_list_select %>% filter(group == "AD_Veh_0"))
comp_samples <- rownames(condition_list_select %>% filter(group == "AD_Veh_10"))

# Perform row-wise t-test
gsva_matrix_4$AD_Veh_0_vs_AD_Veh_10_pvalue <- apply(gsva_matrix_4, 1, function(row) {
  t.test(as.numeric(row[ref_samples]),
         as.numeric(row[comp_samples]),
         paired = FALSE)$p.value
})


# Get column names (sample IDs) for each group
ref_samples <- rownames(condition_list_select %>% filter(group == "CTRL_Veh_0"))
comp_samples <- rownames(condition_list_select %>% filter(group == "AD_Veh_10"))

# Perform row-wise t-test
gsva_matrix_4$CTRL_Veh_0_vs_AD_Veh_10_pvalue <- apply(gsva_matrix_4, 1, function(row) {
  t.test(as.numeric(row[ref_samples]),
         as.numeric(row[comp_samples]),
         paired = FALSE)$p.value
})

write.csv(gsva_matrix_4, "./results/02-GSVA/02_GSVA_matrix_4_group.csv")


gsva_matrix_4_sel <- gsva_matrix_4 %>% filter(CTRL_Veh_0_vs_AD_Veh_0_pvalue < 0.05 & 
                                               AD_Veh_0_vs_AD_Veh_10_pvalue < 0.05 &
                                               CTRL_Veh_0_vs_AD_Veh_10_pvalue >= 0.05)

write.csv(gsva_matrix_4_sel, "./results/02-GSVA/03_GSVA_matrix_4_group_sign.csv")


```


```{r select_pathway, fig.width=4, fig.height=6}

box_plot_folder<- file.path(result_folder,"04-GSVA-4","Boxplot")
# create the folder
dir.create(box_plot_folder, showWarnings = FALSE)

gsva_matrix <- read.csv( "./results/02-GSVA/02_GSVA_matrix_4.csv", row.names = 1)
colnames(gsva_matrix) <- sub("^X", "", colnames(gsva_matrix))
condition_list_label <- condition_list 
condition_list_label$group <- factor(
  condition_list_label$group,
  # levels = c("CTRL_Veh_0", "CTRL_Veh_10",  "CTRL_Veh_100", 
  #            "AD_Veh_0", "AD_Veh_10", "AD_Veh_100")
  levels = c("CTRL_Veh_0", "CTRL_Veh_10",  "AD_Veh_0", "AD_Veh_10")
)

library(forcats)

condition_list_label$group <- fct_recode(condition_list_label$group,
  "CTRL_Veh" = "CTRL_Veh_0",
  "CTRL_10" = "CTRL_Veh_10",
  "AD_Veh" = "AD_Veh_0",
  "AD_10" = "AD_Veh_10"
)

levels(condition_list_label$group) <- c("CTRL_Veh", "CTRL_10", "AD_Veh", "AD_10")






# plot the heatmap for the GSVA result
pathway_list <- rownames(gsva_matrix)
# filter it contain the immune inflame, synap, mitocho, amyloid
pathway_list <- pathway_list[grep("immune|inflame|synap|mitocho|amyloid", pathway_list, ignore.case = TRUE)]

# 
# 
# 
# # plot for the focus pathway
# set.seed(2025)
# for (i in 1:length(pathway_list)){
#   pathway_name <- pathway_list[i]
#   # print(pathway_name)
#   if (i %% 10 == 0){
#     print(paste0("Processing ", i, " pathway"))
#   }
#   p<-plot_gsva_boxplot_mutil_4(gsva_matrix,
#                     condition_list_label =condition_list_label,
#                     pathway_name =  pathway_name,
#                     figure_folder = file.path(result_folder,"02-GSVA-4-group","Boxplot-select"),
#                     file_name = paste0("GSVA_", pathway_name),
#                     fig.height = 6, fig.width = 4, save=TRUE,
#                     reference_group_1="CTRL_Veh" , compare_group_1="CTRL_10",
#                     reference_group_2="AD_Veh"   , compare_group_2="AD_10",
#                     reference_group_3="CTRL_Veh" , compare_group_3="AD_Veh",
#                     reference_group_4="CTRL_10", compare_group_4="AD_10",
#                     reference_group_5="CTRL_Veh" , compare_group_5="AD_10")
#   # print(p)
# }



set.seed(2025)
for (i in 1:10){
  pathway_name <- pathway_list[i]
  # print(pathway_name)
  if (i %% 10 == 0){
    print(paste0("Processing ", i, " pathway"))
  }
  p<-plot_gsva_boxplot_mutil_4(gsva_matrix,
                    condition_list_label =condition_list_label,
                    pathway_name =  pathway_name,
                    figure_folder = file.path(result_folder,"02-GSVA-4-group","Boxplot-select"),
                    file_name = paste0("GSVA_", pathway_name),
                    fig.height = 6, fig.width = 4, save=FALSE,
                    reference_group_1="CTRL_Veh" , compare_group_1="CTRL_10",
                    reference_group_2="AD_Veh"   , compare_group_2="AD_10",
                    reference_group_3="CTRL_Veh" , compare_group_3="AD_Veh",
                    reference_group_4="CTRL_10", compare_group_4="AD_10",
                    reference_group_5="CTRL_Veh" , compare_group_5="AD_10")
  print(p)
}








```


\newpage
## Select the significant pathway


```{r select_pathway_signficant, fig.width=4, fig.height=6}

# plot the heatmap for the GSVA result
pathway_list <- rownames(gsva_matrix_4_sel)
# filter it contain the immune inflame, synap, mitocho, amyloid
pathway_list <- pathway_list[grep("immune|inflame|synap|mitocho|amyloid", pathway_list, ignore.case = TRUE)]


# # plot for the focus pathway
# set.seed(2025)
# for (i in 1:length(pathway_list)){
#   pathway_name <- pathway_list[i]
#   # print(pathway_name)
#   if (i %% 10 == 0){
#     print(paste0("Processing ", i, " pathway"))
#   }
#   p<-plot_gsva_boxplot_mutil_4(gsva_matrix,
#                     condition_list_label =condition_list_label,
#                     pathway_name =  pathway_name,
#                     figure_folder = file.path(result_folder,"02-GSVA-4-group","Boxplot-sign"),
#                     file_name = paste0("GSVA_", pathway_name),
#                     fig.height = 6, fig.width = 4, save=TRUE,
#                     reference_group_1="CTRL_Veh" , compare_group_1="CTRL_10",
#                     reference_group_2="AD_Veh"   , compare_group_2="AD_10",
#                     reference_group_3="CTRL_Veh" , compare_group_3="AD_Veh",
#                     reference_group_4="CTRL_10", compare_group_4="AD_10",
#                     reference_group_5="CTRL_Veh" , compare_group_5="AD_10")
#   # print(p)
# }



# plot for the focus pathway
set.seed(2025)
for (i in 1:10){
  pathway_name <- pathway_list[i]
  # print(pathway_name)
  if (i %% 10 == 0){
    print(paste0("Processing ", i, " pathway"))
  }
  p<-plot_gsva_boxplot_mutil_4(gsva_matrix,
                    condition_list_label =condition_list_label,
                    pathway_name =  pathway_name,
                    figure_folder = file.path(result_folder,"02-GSVA-4-group","Boxplot-sign"),
                    file_name = paste0("GSVA_", pathway_name),
                    fig.height = 6, fig.width = 4, save=FALSE,
                    reference_group_1="CTRL_Veh" , compare_group_1="CTRL_10",
                    reference_group_2="AD_Veh"   , compare_group_2="AD_10",
                    reference_group_3="CTRL_Veh" , compare_group_3="AD_Veh",
                    reference_group_4="CTRL_10", compare_group_4="AD_10",
                    reference_group_5="CTRL_Veh" , compare_group_5="AD_10")
  print(p)
}

```

\newpage

# Session information
```{r}
sessionInfo()
```

