---
title: "GSVA for mutil Group"
author: "Ximing Ran"
date: "2025-03-24"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: false
  html_document:
    # code_folding: hide
    toc: true
    toc_float: true
hitheme: tomorrow
highlighter: highlight.js

---

```{r setup, include=FALSE}
# load libraries
library(tidyverse)
library(knitr)
set.seed(2024)

knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.path = "./results/03-DEG_overlap/Analysis_figure/"
)

theme_set(theme_bw() + theme(legend.position = "bottom"))
knitr::kable(head(mtcars[, 1:4]), "simple")
```

```{r}
library(tibble)
library(tidyr)
library(dplyr)
library(rtracklayer)
```



```{r local_function_load}
# load function from local files
source(here::here("source", "DEG_functions.R"))
```

\newpage

# 1. Read the count data
In this section, we will read the clean count data from the synaptosomes_bulkRNA folder.  We will read the data and merge them into a single table. 

```{r load_bulkRNA_data}
input_count <- read.csv(here::here("data", "bulkRNA",
                                      "bulkRNA_counts_cleaned.csv"))
counts <-  as.data.frame(input_count) %>% 
  column_to_rownames(var = "gene")
colnames(counts) <- gsub("_", "-", colnames(counts))


# raw sample list
sample_list_raw <- read.csv(here::here("data", "bulkRNA",
                                      "sample_info_AD.csv")) %>%
                    mutate(condition = paste0(Diagosis, "_", Treatment),
                           sample = gsub("_", "-", Sample.name))


# Ensure the column names of counts exist in Sample.name
new_colnames <- sample_list_raw$Label[match(colnames(counts), sample_list_raw$sample )]

# Assign new column names
colnames(counts) <- new_colnames


# sort the columns by the colname
condition_list <- data.frame(
  group =sample_list_raw$condition
)

row.names(condition_list) <- sample_list_raw$Label

counts<- counts[, rownames(condition_list)]

gene_name_mapping<- readRDS(here::here("data","ref" ,"gene_name_mapping.rds"))
```



```{r DESeq2_analysis}
# Init the result folder structure for the result
result_folder_all = './results/03-DEG_overlap'
result_folder = result_folder_all

```


\newpage

# 2. Visualization for reuslt

## (1) Sample information - PCA plot

```{r Sample_PCA, fig.width=8, fig.height=6}
figure_folder = result_folder
# do PCA for counts data
dds_obj <- DESeqDataSetFromMatrix(countData = counts,
                                  colData = condition_list,
                                  design = ~ group)
vsd.obj <- varianceStabilizingTransformation(dds_obj, blind = TRUE)
pcaData <- plotPCA(vsd.obj,  intgroup = c("group"), returnData = T)
percentVar <- round(100 * attr(pcaData, "percentVar"))


p <-ggplot(pcaData, aes(PC1, PC2, color=group)) +
  geom_point(size=3) +
  labs(x = paste0("PC1: ",percentVar[1],"% variance"),
       y = paste0("PC2: ",percentVar[2],"% variance"),
  ) +
  stat_ellipse(level = 0.95)+
  theme_bw() +
  # theme_classic()+
  theme(text = element_text(family = "Arial", colour = "black")) +
  # scale_color_manual(values = assigned_colors) +
  ggrepel::geom_text_repel(aes(label = name), color = "black")

print(p)
# ggsave("./results/01-Sample_info/01_sample_PCA_plot.pdf", p,width = 8, height = 6, units = "in", dpi = 300)
# ggsave("./results/01-Sample_info/01_sample_PCA_plot.png", p,width = 8, height = 6, units = "in", dpi = 300)
#   
```

## (2) Sample information - Distance heatmap

```{r Sample_dis_Heatmap, fig.width=8, fig.height=6}
 # Now apply variance stabilizing transformation
 sampleDists <- dist(t(assay(vsd.obj)))
 sampleDistMatrix <- as.matrix( sampleDists )
 rownames(sampleDistMatrix) <- paste( vsd.obj$group )
 colors <- colorRampPalette( rev(RColorBrewer::brewer.pal(9, "Blues")) )(255)
 p <- pheatmap::pheatmap(sampleDistMatrix,
                         clustering_distance_rows = sampleDists,
                         clustering_distance_cols = sampleDists,
                         col = colors) 
print(p)

#  ggsave("./results/01-Sample_info/02_sample_distance_heatmap.pdf", p,width = 8, height = 6, units = "in", dpi = 300)
# ggsave("./results/01-Sample_info/02_sample_distance_heatmap.png",
#        p, width = 8, height = 6, units = "in", dpi = 300)

```

\newpage



# 3. DEG analysis

## (1) LogFC threshold 1



```{r}

logFC_threshold = 1

DEG_AD <- read.csv(here::here("analysis", "02-DEG", 
                              "results", "DEG-V_0",
                              "02-DEG","01_all_gene_results.csv"), row.names = 1)
DEG_AD <- DEG_AD %>% filter(abs(log2FoldChange) > logFC_threshold & padj < 0.05) %>%
          mutate(AD_logFC = log2FoldChange, AD_padj = padj,
                 AD_label = ifelse(log2FoldChange > 0, "up", "down")
                 ) %>%
  
          rownames_to_column("gene") %>%
          dplyr::select(gene, AD_logFC, AD_padj,AD_label)

DEG_AD_drug <- read.csv(here::here("analysis", "02-DEG", 
                              "results", "DEG-AD_0_vs_10",
                              "02-DEG","01_all_gene_results.csv"), row.names = 1)

DEG_AD_drug <- DEG_AD_drug %>% filter(abs(log2FoldChange) > logFC_threshold & padj < 0.05)%>%
          mutate(AD_drug_logFC = log2FoldChange, AD_drug_padj = padj,
                 AD_drug_label = ifelse(log2FoldChange > 0, "up", "down")
                 ) %>%
          rownames_to_column("gene") %>%
          dplyr::select(gene, AD_drug_logFC, AD_drug_padj,AD_drug_label) 

overlap_genes <- intersect(DEG_AD$gene, DEG_AD_drug$gene)

DEG_overlap <- inner_join(DEG_AD %>% filter(gene %in% overlap_genes),
                          DEG_AD_drug %>% filter(gene %in% overlap_genes),
                          by = "gene")

DEG_overlap <- DEG_overlap %>% filter(AD_logFC*AD_drug_logFC < 0)
table(DEG_overlap$AD_label)


# save the result to csv 
write.csv(DEG_overlap, file.path(result_folder,  "DEG_overlap_1.0.csv"), row.names = F)




```


### Heatmap for DEG 

```{r DEG_1_up_heatmap,fig.width=6, fig.height=8}
file_name = "01-DEG_1.0_up"
gene_list <- DEG_overlap %>% filter(AD_label == "up") %>% pull(gene)
condition_list_label <- condition_list 
condition_list_label$group <- factor(
  condition_list_label$group,
  # levels = c("CTRL_Veh_0", "CTRL_Veh_10",  "CTRL_Veh_100", 
  #            "AD_Veh_0", "AD_Veh_10", "AD_Veh_100")
  levels = c("CTRL_Veh_0", "CTRL_Veh_10",  "AD_Veh_0", "AD_Veh_10")
)

library(forcats)

condition_list_label$group <- fct_recode(condition_list_label$group,
  "CTRL_Veh" = "CTRL_Veh_0",
  "CTRL_10" = "CTRL_Veh_10",
  "AD_Veh" = "AD_Veh_0",
  "AD_10" = "AD_Veh_10"
)


# make the color scale
brewer_palette <- "RdBu"
ramp <- colorRampPalette( RColorBrewer::brewer.pal(11, brewer_palette))
mr <- ramp(256)[256:1]

# color map
unique_groups <- c("CTRL_Veh","CTRL_10","AD_Veh","AD_10")
color_palette <- c("#8DD2C5","#BAE3DC",   "#f47f72", "#F6B3AC")
assigned_colors <- setNames(color_palette[seq_along(unique_groups)], unique_groups)
annotation_colors <- list(group = assigned_colors)
annotation_col <- as.data.frame(colData(vsd.obj)[, "group", drop = FALSE])


annotation_col <- annotation_col %>% filter( group %in% c("CTRL_Veh_0", "CTRL_Veh_10",  "AD_Veh_0", "AD_Veh_10"))
annotation_col$group <- fct_recode(annotation_col$group,
  "CTRL_Veh" = "CTRL_Veh_0",
  "CTRL_10" = "CTRL_Veh_10",
  "AD_Veh" = "AD_Veh_0",
  "AD_10" = "AD_Veh_10"
)



stabilized_counts <- assay(vsd.obj)
row_variances <- rowVars(stabilized_counts)

stabilized_counts <- stabilized_counts - rowMeans(stabilized_counts, na.rm=T)
stabilized_counts <- stabilized_counts[gene_list, rownames(annotation_col)]







print(sprintf("Heatmap for %s ", file_name))

# Transparent PNG
png(file.path(figure_folder, "01-DEG_1.0", paste0(file_name, ".png")),
    width = 8, height = 8, units = "in", res = 300, bg = "transparent")
pheatmap::pheatmap(stabilized_counts,
                   color = mr,
                   scale = "none",
                   annotation_col = annotation_col,
                   annotation_colors = annotation_colors,
                   cluster_rows = TRUE,
                   cluster_cols = FALSE,
                   fontsize_col = 10,
                   fontsize_row = 10,
                   treeheight_row = 0,
                   show_rownames = FALSE,
                   border_color = NA)
dev.off()

# Transparent PDF
pdf(file.path(figure_folder, "01-DEG_1.0", paste0(file_name, ".pdf")),
    width = 8, height = 8, bg = "transparent")
pheatmap::pheatmap(stabilized_counts,
                   color = mr,
                   scale = "row",
                   annotation_col = annotation_col,
                   annotation_colors = annotation_colors,
                   cluster_rows = TRUE,
                   cluster_cols = FALSE,
                   fontsize_col = 10,
                   fontsize_row = 10,
                   treeheight_row = 0,
                   show_rownames = FALSE,
                   border_color = NA)
dev.off()




# Transparent PNG
png(file.path(figure_folder, "01-DEG_1.0", paste0(file_name, "_row.png")),
    width = 8, height = 8, units = "in", res = 300, bg = "transparent")
pheatmap::pheatmap(stabilized_counts,
                   color = mr,
                   scale = "row",
                   annotation_col = annotation_col,
                   annotation_colors = annotation_colors,
                   cluster_rows = TRUE,
                   cluster_cols = FALSE,
                   fontsize_col = 10,
                   fontsize_row = 10,
                   treeheight_row = 0,
                   show_rownames = FALSE,
                   border_color = NA)
dev.off()

# Transparent PDF
pdf(file.path(figure_folder, "01-DEG_1.0", paste0(file_name, "_row.pdf")),
    width = 8, height = 8, bg = "transparent")
pheatmap::pheatmap(stabilized_counts,
                   color = mr,
                   scale = "none",
                   annotation_col = annotation_col,
                   annotation_colors = annotation_colors,
                   cluster_rows = TRUE,
                   cluster_cols = FALSE,
                   fontsize_col = 10,
                   fontsize_row = 10,
                   treeheight_row = 0,
                   show_rownames = FALSE,
                   border_color = NA)
dev.off()


```


\newpage

```{r DEG_1_down_heatmap,fig.width=6, fig.height=8}
file_name = "01-DEG_1.0_down"
gene_list <- DEG_overlap %>% filter(AD_label == "down") %>% pull(gene)
condition_list_label <- condition_list 
condition_list_label$group <- factor(
  condition_list_label$group,
  # levels = c("CTRL_Veh_0", "CTRL_Veh_10",  "CTRL_Veh_100", 
  #            "AD_Veh_0", "AD_Veh_10", "AD_Veh_100")
  levels = c("CTRL_Veh_0", "CTRL_Veh_10",  "AD_Veh_0", "AD_Veh_10")
)

library(forcats)

condition_list_label$group <- fct_recode(condition_list_label$group,
  "CTRL_Veh" = "CTRL_Veh_0",
  "CTRL_10" = "CTRL_Veh_10",
  "AD_Veh" = "AD_Veh_0",
  "AD_10" = "AD_Veh_10"
)


# make the color scale
brewer_palette <- "RdBu"
ramp <- colorRampPalette( RColorBrewer::brewer.pal(11, brewer_palette))
mr <- ramp(256)[256:1]

# color map
unique_groups <- c("CTRL_Veh","CTRL_10","AD_Veh","AD_10")
color_palette <- c("#8DD2C5","#BAE3DC",   "#f47f72", "#F6B3AC")
assigned_colors <- setNames(color_palette[seq_along(unique_groups)], unique_groups)
annotation_colors <- list(group = assigned_colors)
annotation_col <- as.data.frame(colData(vsd.obj)[, "group", drop = FALSE])


annotation_col <- annotation_col %>% filter( group %in% c("CTRL_Veh_0", "CTRL_Veh_10",  "AD_Veh_0", "AD_Veh_10"))
annotation_col$group <- fct_recode(annotation_col$group,
  "CTRL_Veh" = "CTRL_Veh_0",
  "CTRL_10" = "CTRL_Veh_10",
  "AD_Veh" = "AD_Veh_0",
  "AD_10" = "AD_Veh_10"
)



stabilized_counts <- assay(vsd.obj)
row_variances <- rowVars(stabilized_counts)

stabilized_counts <- stabilized_counts - rowMeans(stabilized_counts, na.rm=T)
stabilized_counts <- stabilized_counts[gene_list, rownames(annotation_col)]



print(sprintf("Heatmap for %s ", file_name))

# Transparent PNG
png(file.path(figure_folder, "01-DEG_1.0", paste0(file_name, ".png")),
    width = 8, height = 8, units = "in", res = 300, bg = "transparent")
pheatmap::pheatmap(stabilized_counts,
                   color = mr,
                   scale = "none",
                   annotation_col = annotation_col,
                   annotation_colors = annotation_colors,
                   cluster_rows = TRUE,
                   cluster_cols = FALSE,
                   fontsize_col = 10,
                   fontsize_row = 10,
                   treeheight_row = 0,
                   show_rownames = FALSE,
                   border_color = NA)
dev.off()

# Transparent PDF
pdf(file.path(figure_folder, "01-DEG_1.0", paste0(file_name, ".pdf")),
    width = 8, height = 8, bg = "transparent")
pheatmap::pheatmap(stabilized_counts,
                   color = mr,
                   scale = "row",
                   annotation_col = annotation_col,
                   annotation_colors = annotation_colors,
                   cluster_rows = TRUE,
                   cluster_cols = FALSE,
                   fontsize_col = 10,
                   fontsize_row = 10,
                   treeheight_row = 0,
                   show_rownames = FALSE,
                   border_color = NA)
dev.off()




# Transparent PNG
png(file.path(figure_folder, "01-DEG_1.0", paste0(file_name, "_row.png")),
    width = 8, height = 8, units = "in", res = 300, bg = "transparent")
pheatmap::pheatmap(stabilized_counts,
                   color = mr,
                   scale = "row",
                   annotation_col = annotation_col,
                   annotation_colors = annotation_colors,
                   cluster_rows = TRUE,
                   cluster_cols = FALSE,
                   fontsize_col = 10,
                   fontsize_row = 10,
                   treeheight_row = 0,
                   show_rownames = FALSE,
                   border_color = NA)
dev.off()

# Transparent PDF
pdf(file.path(figure_folder, "01-DEG_1.0", paste0(file_name, "_row.pdf")),
    width = 8, height = 8, bg = "transparent")
pheatmap::pheatmap(stabilized_counts,
                   color = mr,
                   scale = "none",
                   annotation_col = annotation_col,
                   annotation_colors = annotation_colors,
                   cluster_rows = TRUE,
                   cluster_cols = FALSE,
                   fontsize_col = 10,
                   fontsize_row = 10,
                   treeheight_row = 0,
                   show_rownames = FALSE,
                   border_color = NA)
dev.off()

       
```


\newpage

### Enrichment analysis for up-regulated genes

```{r Enrichment_DEG_1_up,fig.width=12, fig.height=8}

up_gene_1 <- DEG_overlap %>% filter(AD_label == "up") %>% pull(gene)

# test for the function
Enrichment_analysis(gene_list = up_gene_1,
                    result_folder = file.path(result_folder,"01-DEG_1.0"),
                    file_name = "01-DEG_1.0_up", gene_name_mapping, flag = "Up")

```
\newpage

### Enrichment analysis for down-regulated genes

```{r Enrichment_DEG_1_down,fig.width=12, fig.height=8}
down_gene_1 <- DEG_overlap %>% filter(AD_label == "down") %>% pull(gene)

# test for the function
Enrichment_analysis(gene_list = down_gene_1,
                    result_folder = file.path(result_folder,"01-DEG_1.0"),
                    file_name = "01-DEG_1.0_down", gene_name_mapping, flag = "Down")

```
\newpage

### Enrichment plot for selected pathway

```{r Enrichment_DEG_1_down_select,fig.width=12, fig.height=8}

result_GOBP <- read.csv(here::here("data", "ref", "DEG_1.0_down_GO_BP_WN_2025-04-01.csv"))


flag = "Down"

result_GOBP 


result_GOBP <- result_GOBP %>%
  dplyr::select(term_name, p_value) %>%
  arrange(p_value) %>%
  mutate(term_name = factor(term_name, levels = rev(unique(term_name))))
# Select the top 30 smallest p-values

p <- ggplot(result_GOBP , aes(x = term_name, y = -log10(p_value), fill = -log10(p_value))) +
  geom_bar(stat = "identity", show.legend = TRUE) +
  {
    if (flag == "Up") {
      scale_fill_gradient(high = "#a50f15", low = "#fc9272")  # Red scale for UP
    } else if (flag == "Down") {
      scale_fill_gradient(low = "#56B1F7", high = "#132B43")  # Blue scale for DOWN
    } else {
      scale_fill_manual(values = rep("lightblue", nrow(result_df_30)))  # Default to light blue
    }
  } +
  ylab("-log10(FDR)") +
  xlab("") +
  ggtitle("Select Top Pathways for GOBP (Sorted by FDR)") +
  coord_flip() +
  theme_classic(base_size = 14) +
  theme(
    text = element_text(family = "Arial", size = 14, colour = "black"),
    axis.title = element_text(color = "black"),
    axis.text = element_text(color = "black"),
    plot.title = element_text(family = "Arial", size = 14, face = "bold", hjust = 0.5),
    panel.background = element_rect(fill = "transparent", color = NA),  # Transparent panel
    plot.background = element_rect(fill = "transparent", color = NA),   # Transparent plot
    legend.background = element_rect(fill = "transparent", color = NA), # Transparent legend background
    legend.key = element_rect(fill = "transparent", color = NA)         # Transparent legend key box
  ) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "black")

print(p)

ggsave(file.path(result_folder, "01-DEG_1.0", "01-DEG_1.0_down_GOBP_select.png"), 
       p, width = 12, height = 8, units = "in", dpi = 300, bg = "transparent")

ggsave(file.path(result_folder, "01-DEG_1.0", "01-DEG_1.0_down_GOBP_select.pdf"), 
       p, width = 12, height = 8, units = "in", bg = "transparent")

  
```


\newpage
```{r Enrichment_DEG_1_up_select,fig.width=12, fig.height=8}

result_GOBP <- read.csv(file.path(result_folder, "01-DEG_1.0", "01-DEG_1.0_up.csv"))


flag = "Up"

result_GOBP 


result_GOBP <- result_GOBP %>%
  dplyr::select(term_name, p_value) %>%
  arrange(p_value) %>%
  
    slice_head(n = 30) %>%
  mutate(term_name = factor(term_name, levels = rev(unique(term_name))))
# Select the top 30 smallest p-values

p <- ggplot(result_GOBP , aes(x = term_name, y = -log10(p_value), fill = -log10(p_value))) +
  geom_bar(stat = "identity", show.legend = TRUE) +
  {
    if (flag == "Up") {
      scale_fill_gradient(high = "#a50f15", low = "#fc9272")  # Red scale for UP
    } else if (flag == "Down") {
      scale_fill_gradient(low = "#56B1F7", high = "#132B43")  # Blue scale for DOWN
    } else {
      scale_fill_manual(values = rep("lightblue", nrow(result_df_30)))  # Default to light blue
    }
  } +
  ylab("-log10(FDR)") +
  xlab("") +
  ggtitle("Top 30 Pathways (Sorted by FDR)") +
  coord_flip() +
  theme_classic(base_size = 14) +
  theme(
    text = element_text(family = "Arial", size = 14, colour = "black"),
    axis.title = element_text(color = "black"),
    axis.text = element_text(color = "black"),
    plot.title = element_text(family = "Arial", size = 14, face = "bold", hjust = 0.5),
    panel.background = element_rect(fill = "transparent", color = NA),  # Transparent panel
    plot.background = element_rect(fill = "transparent", color = NA),   # Transparent plot
    legend.background = element_rect(fill = "transparent", color = NA), # Transparent legend background
    legend.key = element_rect(fill = "transparent", color = NA)         # Transparent legend key box
  ) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "black")

print(p)

ggsave(file.path(result_folder, "01-DEG_1.0", "01-DEG_1.0_up_select.png"), 
       p, width = 12, height = 8, units = "in", dpi = 300, bg = "transparent")

ggsave(file.path(result_folder, "01-DEG_1.0", "01-DEG_1.0_up_select.pdf"), 
       p, width = 12, height = 8, units = "in", bg = "transparent")

  
```




\newpage

## (2) LogFC threshold 1.5
```{r}

logFC_threshold = 1.5
DEG_AD <- read.csv(here::here("analysis", "02-DEG", 
                              "results", "DEG-V_0",
                              "02-DEG","01_all_gene_results.csv"), row.names = 1)
DEG_AD <- DEG_AD %>% filter(abs(log2FoldChange) > logFC_threshold & padj < 0.05) %>%
          mutate(AD_logFC = log2FoldChange, AD_padj = padj,
                 AD_label = ifelse(log2FoldChange > 0, "up", "down")
                 ) %>%
  
          rownames_to_column("gene") %>%
          dplyr::select(gene, AD_logFC, AD_padj,AD_label)

DEG_AD_drug <- read.csv(here::here("analysis", "02-DEG", 
                              "results", "DEG-AD_0_vs_10",
                              "02-DEG","01_all_gene_results.csv"), row.names = 1)

DEG_AD_drug <- DEG_AD_drug %>% filter(abs(log2FoldChange) > logFC_threshold & padj < 0.05)%>%
          mutate(AD_drug_logFC = log2FoldChange, AD_drug_padj = padj,
                 AD_drug_label = ifelse(log2FoldChange > 0, "up", "down")
                 ) %>%
          rownames_to_column("gene") %>%
          dplyr::select(gene, AD_drug_logFC, AD_drug_padj,AD_drug_label) 

overlap_genes <- intersect(DEG_AD$gene, DEG_AD_drug$gene)

DEG_overlap <- inner_join(DEG_AD %>% filter(gene %in% overlap_genes),
                          DEG_AD_drug %>% filter(gene %in% overlap_genes),
                          by = "gene")

DEG_overlap <- DEG_overlap %>% filter(AD_logFC*AD_drug_logFC < 0)
table(DEG_overlap$AD_label)


# save the result to csv 
write.csv(DEG_overlap, file.path(result_folder,  "DEG_overlap_1.5.csv"), row.names = F)




```




### Heatmap for DEG 

```{r DEG_1.5_up_heatmap,fig.width=6, fig.height=8}
file_name = "01-DEG_1.5_up"
gene_list <- DEG_overlap %>% filter(AD_label == "up") %>% pull(gene)
condition_list_label <- condition_list 
condition_list_label$group <- factor(
  condition_list_label$group,
  # levels = c("CTRL_Veh_0", "CTRL_Veh_10",  "CTRL_Veh_100", 
  #            "AD_Veh_0", "AD_Veh_10", "AD_Veh_100")
  levels = c("CTRL_Veh_0", "CTRL_Veh_10",  "AD_Veh_0", "AD_Veh_10")
)

library(forcats)

condition_list_label$group <- fct_recode(condition_list_label$group,
  "CTRL_Veh" = "CTRL_Veh_0",
  "CTRL_10" = "CTRL_Veh_10",
  "AD_Veh" = "AD_Veh_0",
  "AD_10" = "AD_Veh_10"
)


# make the color scale
brewer_palette <- "RdBu"
ramp <- colorRampPalette( RColorBrewer::brewer.pal(11, brewer_palette))
mr <- ramp(256)[256:1]

# color map
unique_groups <- c("CTRL_Veh","CTRL_10","AD_Veh","AD_10")
color_palette <- c("#8DD2C5","#BAE3DC",   "#f47f72", "#F6B3AC")
assigned_colors <- setNames(color_palette[seq_along(unique_groups)], unique_groups)
annotation_colors <- list(group = assigned_colors)
annotation_col <- as.data.frame(colData(vsd.obj)[, "group", drop = FALSE])


annotation_col <- annotation_col %>% filter( group %in% c("CTRL_Veh_0", "CTRL_Veh_10",  "AD_Veh_0", "AD_Veh_10"))
annotation_col$group <- fct_recode(annotation_col$group,
  "CTRL_Veh" = "CTRL_Veh_0",
  "CTRL_10" = "CTRL_Veh_10",
  "AD_Veh" = "AD_Veh_0",
  "AD_10" = "AD_Veh_10"
)



stabilized_counts <- assay(vsd.obj)
row_variances <- rowVars(stabilized_counts)

stabilized_counts <- stabilized_counts - rowMeans(stabilized_counts, na.rm=T)
stabilized_counts <- stabilized_counts[gene_list, rownames(annotation_col)]



print(sprintf("Heatmap for %s ", file_name))

p <-pheatmap::pheatmap(stabilized_counts,
                         color = mr,
                         scale = "none",
                         annotation_col = annotation_col,
                         annotation_colors = annotation_colors,
                         cluster_rows = TRUE,
                         cluster_cols = FALSE ,
                         fontsize_col = 10,
                         fontsize_row = 10,
                         treeheight_row = 0,
                         show_rownames = FALSE,
                         border_color = NA)
  
ggsave(file.path(figure_folder, "02-DEG_1.5", paste0(file_name, ".png")), 
       p, width = 8, height = 8, units = "in", dpi = 300)
ggsave(file.path(figure_folder, "02-DEG_1.5", paste0(file_name, ".pdf")),
       p, width = 8, height = 8, units = "in")


p <-pheatmap::pheatmap(stabilized_counts,
                         color = mr,
                         scale = "row",
                         annotation_col = annotation_col,
                         annotation_colors = annotation_colors,
                         cluster_rows = TRUE,
                         cluster_cols = FALSE ,
                         fontsize_col = 10,
                         fontsize_row = 10,
                         treeheight_row = 0,
                         show_rownames = FALSE,
                         border_color = NA)
  
ggsave(file.path(figure_folder, "02-DEG_1.5", paste0(file_name, "_row.png")), 
       p, width = 8, height = 8, units = "in", dpi = 300)
ggsave(file.path(figure_folder, "02-DEG_1.5", paste0(file_name, "_row.pdf")),
       p, width = 8, height = 8, units = "in")
```


\newpage

```{r DEG_1.5_down_heatmap,fig.width=6, fig.height=8}
file_name = "01-DEG_1.5_down"
gene_list <- DEG_overlap %>% filter(AD_label == "down") %>% pull(gene)
condition_list_label <- condition_list 
condition_list_label$group <- factor(
  condition_list_label$group,
  # levels = c("CTRL_Veh_0", "CTRL_Veh_10",  "CTRL_Veh_100", 
  #            "AD_Veh_0", "AD_Veh_10", "AD_Veh_100")
  levels = c("CTRL_Veh_0", "CTRL_Veh_10",  "AD_Veh_0", "AD_Veh_10")
)

library(forcats)

condition_list_label$group <- fct_recode(condition_list_label$group,
  "CTRL_Veh" = "CTRL_Veh_0",
  "CTRL_10" = "CTRL_Veh_10",
  "AD_Veh" = "AD_Veh_0",
  "AD_10" = "AD_Veh_10"
)


# make the color scale
brewer_palette <- "RdBu"
ramp <- colorRampPalette( RColorBrewer::brewer.pal(11, brewer_palette))
mr <- ramp(256)[256:1]

# color map
unique_groups <- c("CTRL_Veh","CTRL_10","AD_Veh","AD_10")
color_palette <- c("#8DD2C5","#BAE3DC",   "#f47f72", "#F6B3AC")
assigned_colors <- setNames(color_palette[seq_along(unique_groups)], unique_groups)
annotation_colors <- list(group = assigned_colors)
annotation_col <- as.data.frame(colData(vsd.obj)[, "group", drop = FALSE])


annotation_col <- annotation_col %>% filter( group %in% c("CTRL_Veh_0", "CTRL_Veh_10",  "AD_Veh_0", "AD_Veh_10"))
annotation_col$group <- fct_recode(annotation_col$group,
  "CTRL_Veh" = "CTRL_Veh_0",
  "CTRL_10" = "CTRL_Veh_10",
  "AD_Veh" = "AD_Veh_0",
  "AD_10" = "AD_Veh_10"
)



stabilized_counts <- assay(vsd.obj)
row_variances <- rowVars(stabilized_counts)

stabilized_counts <- stabilized_counts - rowMeans(stabilized_counts, na.rm=T)
stabilized_counts <- stabilized_counts[gene_list, rownames(annotation_col)]



print(sprintf("Heatmap for %s ", file_name))

p <-pheatmap::pheatmap(stabilized_counts,
                         color = mr,
                         scale = "none",
                         annotation_col = annotation_col,
                         annotation_colors = annotation_colors,
                         cluster_rows = TRUE,
                         cluster_cols = FALSE ,
                         fontsize_col = 10,
                         fontsize_row = 10,
                         treeheight_row = 0,
                         show_rownames = FALSE,
                         border_color = NA)
  
ggsave(file.path(figure_folder, "02-DEG_1.5", paste0(file_name, ".png")), 
       p, width = 8, height = 8, units = "in", dpi = 300)
ggsave(file.path(figure_folder, "02-DEG_1.5", paste0(file_name, ".pdf")),
       p, width = 8, height = 8, units = "in")


p <-pheatmap::pheatmap(stabilized_counts,
                         color = mr,
                         scale = "row",
                         annotation_col = annotation_col,
                         annotation_colors = annotation_colors,
                         cluster_rows = TRUE,
                         cluster_cols = FALSE ,
                         fontsize_col = 10,
                         fontsize_row = 10,
                         treeheight_row = 0,
                         show_rownames = FALSE,
                         border_color = NA)
  
ggsave(file.path(figure_folder, "02-DEG_1.5", paste0(file_name, "_row.png")), 
       p, width = 8, height = 8, units = "in", dpi = 300)
ggsave(file.path(figure_folder, "02-DEG_1.5", paste0(file_name, "_row.pdf")),
       p, width = 8, height = 8, units = "in")
       
```

\newpage

### Enrichment analysis for up-regulated genes

```{r Enrichment_DEG_1.5_up,fig.width=12, fig.height=8}

up_gene_1.5 <- DEG_overlap %>% filter(AD_label == "up") %>% pull(gene)

# test for the function
Enrichment_analysis(gene_list = up_gene_1.5,
                    result_folder = file.path(result_folder,"02-DEG_1.5"),
                    file_name = "01-DEG_1.5_up", gene_name_mapping, flag = "Up")

```

\newpage

### Enrichment analysis for down-regulated genes

```{r Enrichment_DEG_1.5_down,fig.width=12, fig.height=8}
down_gene_1.5 <- DEG_overlap %>% filter(AD_label == "down") %>% pull(gene)

# test for the function
Enrichment_analysis(gene_list = down_gene_1.5,
                    result_folder = file.path(result_folder,"02-DEG_1.5"),
                    file_name = "01-DEG_1.5_down", gene_name_mapping, flag = "Down")

```


\newpage
# Session information
```{r}
sessionInfo()
```

